---
title: "Task 2"
author: "GEOG-490R"
date: 'Spring 2024 - Week 2'
output: html_document
---
## Exploring a dataset

Here is your first task. It should be very straightforward. 

You are going to create a simple plot, and answer a few questions, with a built-in dataset contained in R. You will be using base functions in R, which is to say that you do not need to load any additional libraries. *We'll learn several new, useful libraries throughout the semester.*

**For this task** you will recreate the following graph:

```{r, echo=FALSE}
# Load built-in mtcars dataset
data(mtcars)

# Create scatterplot using plot()
plot(mtcars$wt, 
     mtcars$mpg, 
     main="Motor Vehicle Mileage vs Weight",
     xlab="Weight",
     ylab="Miles Per Gallon")

# Add regression line 
abline(lm(mtcars$mpg ~ mtcars$wt))
```

This code uses the `mtcars` dataset, this data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

The code above creates a basic scatterplot of weight (wt) vs miles per gallon (mpg), adds axis labels and a title, and plots a regression line. 

The key functions from base R are:

- `data()` to load a built-in dataset
- `plot()` to create the initial scatterplot
- `abline()` to add a regression line
- `lm()` to fit linear model for the regression line

To learn more about any function you can type in the function name, for example `plot` will show you the basic function. For a more detailed explanation of a function, type `?plot` to open the help window in the Rstudio console.

### Data description

In addition to recreating the plot above, you will need to describe the dataset we are using. There are many different kinds of data that can be used

```{r, echo=FALSE}
# First few rows 
# head(mtcars, 3)
print("...what is this?")
```

Here are some useful R functions to investigate a dataframe in R markdown, along with brief explanations:

- `class()`: View kind of R object for any variable
- `colnames()`: View column names of dataframe
- `dim()`: View the dimension, or number of rows and columns 
- `summary()`: Summary stats for each column
- `head()/tail()`: View first/last few rows
- several other functions exist as well

### Your task

For this assignment, you will create an R script that describes the `mtcars` dataset using the functions above. You will also write a brief paragraph that summarizes this dataset and results from the plots you created.

#### Submitting

Upload your script to our GitHub classroom with the following link:

[GitHub Classroom Assignment](https://classroom.github.com/a/T_DG18zP)


*Please reach out if you have any questions or concerns!*

**This assignment is due in one week but may be turned for an additional week with a late penalty of -5%**
