---
title: "Practice"
author: "GEOG-490R"
date: '2024-01-24'
output: html_document
---


Use the `mtcars` dataset to understand basic relationships between variables. 

- The relationship between horsepower and fuel efficiency (mpg). A scatterplot can show if they are linearly related. Is engine power related to gas mileage?

- The distribution of fuel efficiency (mpg). A histogram can show the overall spread and central tendency. How is gas mileage distributed across cars?

- Differences in fuel efficiency by number of cylinders. A grouped boxplot can highlight differences in mpg by the variable cyl. Do cars with more cylinders have lower average mpg?

- The relationship between weight and fuel efficiency. A scatterplot can show if heavier cars tend to have lower mpg. Does weight impact gas mileage?

- Differences in fuel efficiency by transmission type. Group the mpg by am (transmission) and compare with a boxplot or densities plot. Do automatic or manual transmissions have better mpg?

- Horsepower differences by number of cylinders. Use a violin plot or boxplot to compare hp by cyl. Do cars with more cylinders tend to have higher horsepower?

- Weight differences by number of cylinders. Use a grouped boxplot to see how weight varies by cyl. Do cars with more cylinders tend to be heavier?


- Calculate the correlation between mpg and wt. Is this correlation statistically significant at the 0.05 level?

- Using the mtcars data, test whether the average mpg is different between cars with 4, 6, and 8 cylinders using ANOVA

- Write a function that simulates 10 coin flips and returns the number of heads. Run this simulation 1000 times and plot the distribution of heads.

- The relationship between horsepower and fuel efficiency (mpg). A scatterplot can show if they are linearly related. Is engine power related to gas mileage?

- The distribution of fuel efficiency (mpg). A histogram can show the overall spread and central tendency. How is gas mileage distributed across cars?

- Differences in fuel efficiency by number of cylinders. A grouped boxplot can highlight differences in mpg by the variable cyl. Do cars with more cylinders have lower average mpg?

- The relationship between weight and fuel efficiency. A scatterplot can show if heavier cars tend to have lower mpg. Does weight impact gas mileage?

- Differences in fuel efficiency by transmission type. Group the mpg by am (transmission) and compare with a boxplot or densities plot. Do automatic or manual transmissions have better mpg?

- Horsepower differences by number of cylinders. Use a violin plot or boxplot to compare hp by cyl. Do cars with more cylinders tend to have higher horsepower?

- Weight differences by number of cylinders. Use a grouped boxplot to see how weight varies by cyl. Do cars with more cylinders tend to be heavier?



 Here is the R markdown with the questions inserted above each code chunk:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
data(mtcars)
```

**The relationship between horsepower and fuel efficiency (mpg)**

```{r}
# The relationship between horsepower and fuel efficiency (mpg). A scatterplot can show if they are linearly related. Is engine power related to gas mileage?

ggplot(mtcars, aes(x=hp, y=mpg)) +
  geom_point() +
  geom_smooth(method='lm')
```

**The distribution of fuel efficiency (mpg)**

```{r}
# The distribution of fuel efficiency (mpg). A histogram can show the overall spread and central tendency. How is gas mileage distributed across cars?

ggplot(mtcars, aes(mpg)) +
  geom_histogram(bins=10)
```

**Differences in fuel efficiency by number of cylinders**

```{r}
# Differences in fuel efficiency by number of cylinders. A grouped boxplot can highlight differences in mpg by the variable cyl. Do cars with more cylinders have lower average mpg?

ggplot(mtcars, aes(x=as.factor(cyl), y=mpg)) +
  geom_boxplot()
```

**The relationship between weight and fuel efficiency** 

```{r}
# The relationship between weight and fuel efficiency. A scatterplot can show if heavier cars tend to have lower mpg. Does weight impact gas mileage?

ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point() +
  geom_smooth(method='lm')
```

**Differences in fuel efficiency by transmission type**

```{r}  
# Differences in fuel efficiency by transmission type. Group the mpg by am (transmission) and compare with a boxplot or densities plot. Do automatic or manual transmissions have better mpg?

ggplot(mtcars, aes(x=am, y=mpg)) +
  geom_boxplot()
```

**Horsepower differences by number of cylinders**

```{r}
# Horsepower differences by number of cylinders. Use a violin plot or boxplot to compare hp by cyl. Do cars with more cylinders tend to have higher horsepower?

ggplot(mtcars, aes(x=as.factor(cyl), y=hp)) +
  geom_violin()
```

**Weight differences by number of cylinders**

```{r}
# Weight differences by number of cylinders. Use a grouped boxplot to see how weight varies by cyl. Do cars with more cylinders tend to be heavier?

ggplot(mtcars, aes(x=as.factor(cyl), y=wt)) +
  geom_boxplot()
```

**Correlation between mpg and wt**

```{r}
# Calculate the correlation between mpg and wt. Is this correlation statistically significant at the 0.05 level?

cor.test(mtcars$mpg, mtcars$wt)
```

**ANOVA test for mpg differences by cylinder**

```{r}  
# Using the mtcars data, test whether the average mpg is different between cars with 4, 6, and 8 cylinders using ANOVA

cyl_groups <- split(mtcars$mpg, mtcars$cyl)
aov_result <- aov(mpg ~ cyl, data=mtcars)
summary(aov_result)
```

**Simulation of coin flips**

```{r}
# Write a function that simulates 10 coin flips and returns the number of heads. Run this simulation 1000 times and plot the distribution of heads.

flip_coin <- function() {
  heads <- sum(sample(c(0,1), 10, replace=TRUE))
  return(heads)
}

sims <- replicate(1000, flip_coin())  
ggplot(data.frame(sims), aes(sims)) +
  geom_histogram(bins=11)
```