---
title: "Final Project"
author: "GEOG-490R"
date: "Spring 2024"
output: html_document
---


### Final Project Template

Your final project should be organized into a cohesive narrative. It does not need to be  long or exhaustive. Include important information, code and figures.

\

#### 1. Question and Data

Define your question, cite all datasets, and describe your data using R functions. Provide a statement that clearly defines your objective or plan for the project.

Any background information can go here.

\

#### 2. Analysis

This section should show your data preparation (cleaning/modifying) and analysis. Justify any steps you took and show your code.

```{r}
print("Show your work!")
```

- a list might be useful

You should include quantitative results in this or the next section.

\

#### 3. Results

This section will include your final maps, figures, and conclusion.

```{r}
v = 1:10
plot(v**4~v, main="Fourth power plot")
```

Your maps and figures don't need to be perfect, but put some time into creating each. Consider balance, color, and other aesthetics.

Ask questions if you are not sure about the expectations.

\